---
title: "README"
author: "Daniel Enns"
date: "2024-10-17"
output:
  github_document: default
always_allow_html: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(warning = FALSE)
```

# Study outline

Freshwater ecosystems exists in very heterogenous landscapes, including not only natural, but also anthropogenic "spatial features"[^1]. Can these features, like barriers, wastewater treatment plants, road & railroad crossings, etc. explain the ecological status of the freshwater fauna? In this study, we try to answer this question combining various spatial data sets and utilizing machine learning algorithms[^2].

[^1]: [Carlisle et al. 2009](https://doi.org/10.1007/s10661-008-0256-z); [Marzin et al. 2013](https://doi.org/10.1007/s10750-012-1254-2); [Murphy & Davy-Bowker 2005](https://doi.org/10.1007/s10750-004-1451-8)

[^2]: [Dedman et al. 2015](http://dx.doi.org/10.1016/j.ecolmodel.2015.05.010); [Elith et al. 2008](https://doi.org/10.1111/j.1365-2656.2008.01390.x); [Heß et al. 2023](http://dx.doi.org/10.1016/j.scitotenv.2023.162952); [Yu et al. 2020](https://doi.org/10.1016/j.ecolmodel.2020.109202)

# Methods

```{r load_libraries_data, include=FALSE}
library(readxl)
library(tidyverse)
library(sf)
library(sfnetworks)
library(SSN2)
library(SSNbler)
library(riverdist)
library(rivernet)
library(geodata)
library(whitebox)
library(osmextract)
library(httr2)
library(tmap)

stream_net <- st_read("./Data/stream_net/Gewässerstrukturgüte_Hessen_rev.shp") %>% 
  select(geometry, GESAMT, HP3, GEWKZ, ABS) %>% st_cast("LINESTRING") %>% 
  mutate(ABS = as.numeric(ABS)) 
```

## Combining Spatial Data

For the spatial analysis all necessary shape and raster files are projected into EPSG . The HLNUG stream network is a collection of 100 m long line segments, each containing a stream ID and a segment ID. The stream ID is designed in such a way, that it represents the one from the subsequent stream in which the current flows into, followed by additional numbers. The segment number increases towards the source of a stream (Fig. 1). Each 100 m segment already contains information on the habitats structural quality (1 - pristine, 7 - completly altered).

```{r create_figure,echo=FALSE , fig.cap = "Fig. 1: Example for stream- and segment ID. The segment number increases in the direction of the source. The stream ID of smaller confluence resembels always that of the bigger stream, in which it flows into, plus additional numbers."}


str_net_plot <- stream_net %>% filter(str_starts(GEWKZ, "24848292")) %>% 
  ggplot()+
  geom_sf(
    aes(
      col = GEWKZ
    ),
    linewidth = 4
  )+
  geom_sf_text(
    aes(label = ABS)
    )+
  labs(col = "stream ID")+
  theme_bw()+
  theme(
    axis.title = element_blank(),
    legend.position = "bottom"
  )
str_net_plot
```

The stream network can be dissolved in QGIS using the dissolve tool and subsequently we can build an initial landscape network (LSN) using the `lines_to_lsn()` function from the [SSNbler](https://cran.r-project.org/web/packages/SSNbler/index.html) package, where we set the snap tolerance of nodes to 3 m and the topology tolerance to 30 m (note that the digitized flow needs to lead from source to outlet points). The summary of the node error report reveals that the network is highly erroneous. (For more detail on the LSN parameters and error types see [Peterson & Pearse 2024](https://github.com/pet221/SSNbler/blob/main/inst/tutorials/Topology_Editing/QGIS/TopologyEditing_QGIS.pdf))

```{r read_node_errors, include=FALSE}
node_errors <- st_read("./Data/SSNbler/node_errors.gpkg") %>% modify_if(is.character, as.factor)
```

```{r summary_node_errors, echo=FALSE}
summary(node_errors)
```

Fixing topological errors in the entire network, manually and using GRASS GIS tools, would take too much time.
