---
title: "README"
author: 
  - name: Daniel Enns
    orcid: 0000-0003-1742-8743
    email: daniel.enns@stud.uni-frankfurt.de
date: "2024-10-17"
format: 
  html:
    toc: true
    toc-location: left
    number-sections: true
    bibliography: references.bib
    code-copy: true
    execute: 
      echo: false
      warning: false
      message: false
    fig-cap-location: bottom
    tbl-cap-location: top
---

# Study outline

Freshwater ecosystems exists in very heterogenous landscapes wich includes not only natural, but also anthropogenic "spatial features"[@carlisle2008; @marzin2012; @murphy2005]. These features can influence freshwater ecosystems and its inhabitants. This gives rise to the question if these features, like barriers, wastewater treatment plants (WWTPs), road & railroad crossings, etc. can explain the ecological status of the freshwater fauna? In this study, we try to answer this question combining various spatial data sets and utilizing machine learning algorithms [@dedman2015; @elith2008; @heß2023; @yu2020].

# Methods

```{r}
#| label: load_libraries_data

library(readxl)
library(tidyverse)
library(stars)
library(sf)
library(sfnetworks)
library(spdep)
library(tidygraph)
library(geodata)
library(whitebox)
library(osmextract)
library(httr2)
library(tmap)
library(ggpubr)
library(gridExtra)
library(PerformanceAnalytics)

setwd("C:/Users/Daniel Enns/Documents/Promotion/MZB/Enns-et-al.-3/Data")
load("Data.RData")

```

## Library

The following packages are needed for the analysis:

``` r
# general
library(readxl)
library(tidyverse)
library(doParallel)

# spatial analysis
library(stars)
library(sf)
library(sfnetworks)
library(tidygraph)
library(nngeo)
library(geodata)
library(whitebox)
library(osmextract)

# modeling
library(caret)
library(mgcv)
library(xgboost)
library(pdp)
```

## Combining Spatial Data

### Point features

For the spatial analysis all necessary shape and raster files are projected into EPSG 25832.

The freshwater invertebrate sampling sites, as well as the data on WWTPs, stormwater overflows (SWOFs) and barriers can readily be loaded into R. Sampling sites located on large Rivers (e.g. Rhine, Main, Weser) were excluded since data on the complete catchment was not available.

``` r
# load and filter invertebrate sampling sites
mzb <- read_delim("./MZB.csv", delim = ";") %>% 
  bind_cols(mzb, stream_net[st_nearest_feature(mzb, stream_net),]) %>% 
  select(-geom) %>% 
  filter(!GEWKZ  %in% c("2", "238", "24", "41", "4", "44", "239152", "23932", "2396", "23988", "24779892"))
```

Highway and railroad networks can be read in from OpenStreetMap pbf files via the `osmextract` package. These can be intersected with the stream network to create point features. In the osm highway and railroad networks, each lane / track is represented as an individual line feature, so that multiple crossings are present in locations with high lane / track density.

``` r
# Extract highway and railroad networks
transport_net <- oe_read("./hessen-latest.osm.pbf", extra_tags = "railway") %>%
  filter(highway == "motorway" | railway == "rail") %>% 
  st_transform(st_crs(stream_net))

# Intersect with stream network
crossings <- st_intersection(stream_net, transport_net) %>% 
  st_cast("POINT")
```

### Watershed delineation

In order to delineate upstream watersheds for each bio sampling site the SRTM GL1 30m [@opentopography2013] digital elevation model will be used to extract flow accumulation and pointer grids, from which the watersheds are calculated. However, the delineation algorithms struggles in regions with low profiles (namely the Rhine basin), creating an inaccurate stream raster and watersheds. To improve accuracy, the existing stream network is burned into the DEM following these steps in QGIS:

1.  Reproject raster network to stream CRS (EPSG: 25832)

2.  dissolve the stream network

3.  Run the Grass r.carve algorithm with a stream width of 60 m, depth of 3 m and with no flat areas in direction of stream flow allowed (Note: the algorithm does not work on latitude-longitude CRS, see [r.carve GRASS GIS manual](https://grass.osgeo.org/grass-stable/manuals/r.carve.html))

The algorithm will take some time to run (approximately 1 hour and 18 minutes with an Intel core i7, 16 GB RAM). Finally, watersheds can be extracted using the `whitebox` package, following the instructions from [Gannon, 2024](https://vt-hydroinformatics.github.io/Quarto_Book/15-Watershed-Delineation-and-Analysis.html).

``` r
# Breach and fill pits in raster
wbt_breach_depressions_least_cost(
  dem = "./SRTMGL1_30m_2px_burned.tif", 
  output = "./srtm_breach.tif", 
  dist = 7,
  fill = T
)

# Create flow accumulation & pointer grids
wbt_d8_flow_accumulation(
  input = "./srtm_breach.tif",
  output = "./D8FA.tif"
)

wbt_d8_pointer(
  dem = "./srtm_breach.tif",
  output = "./D8pointer.tif"
)

# Extract streams
wbt_extract_streams(
  flow_accum = "./D8FA.tif",
  output = "./stream_raster.tif",
  threshold = 700
)

# Snap points to stream raster
wbt_jenson_snap_pour_points(
  pour_pts = "./MZB.shp",
  streams = "./stream_raster.tif",
  output = "./MZB_snap.shp",
  snap_dist = 300
)

# Delineate watersheds
wbt_watershed(
  d8_pntr = "./D8pointer.tif",
  pour_pts = "./MZB_snap.shp",
  output = "./watersheds.tif"
)
```

Figure 1 demonstrates the structure of the watersheds, which were build consecutively between multiple pour points.

```{r}
#| label: watershed_example_figure
#| fig-cap: "Figure 1: Example watershed. Each watershed is build consecutively between specified pourpoints."

ws_stream <- stream_net %>% filter(str_starts(GEWKZ, "25872")) 
ws_bio <- mzb %>% filter(str_starts(GEWKZ, "25872"))
ws_clip <- ws %>% st_filter(ws_stream, predicate = st_touches) %>% filter(watersheds.tif != 812)

ggplot()+
  geom_sf(
    ws_clip,
    mapping = aes(fill = as.factor(watersheds.tif))
  )+
  geom_sf(
    ws_stream,
    mapping = aes(),
    linewidth = 1
  )+
  geom_sf(
    ws_bio,
    mapping = aes(),
    fill = "#FFFF00",
    colour = "#000000",
    size = 2,
    shape = 21
  )+
  scale_x_continuous(breaks = c(8.08, 8.12, 8.16, 8.20))+
  theme_bw()+
  theme(
    legend.position = "none"
  )
```

The created watershed raster can be read in and converted to polygon features, which subsequently be used to intersect the [CORINE Land Cover](https://land.copernicus.eu/en/products/corine-land-cover) polygons. The area of each intersection can be calculated using the `st_area()` function.

``` r
# Read in Watersheds and convert to polygon 
ws_poly <- read_stars("./watersheds.tif") %>% 
  st_as_sf(as_points = F, merge = T)

# Create dataframe with land cover areas
landcover_ws <- st_intersection(landcover, ws_poly) %>% 
  mutate(area = st_area(.)) %>% st_drop_geometry() %>% 
  as.data.frame() %>% filter(!is.na(type))

# Summarize land cover types in each watershed
landcover_sum <- landcover_ws %>% group_by(watersheds.tif, type) %>% 
  summarize(area = sum(area)) %>% 
  pivot_wider(names_from = type, values_from = area)

# Combine watersheds with summarized land cover
watersheds_lc <- left_join(ws_poly, landcover_sum, by = "watersheds.tif")
```

### Network building

The Hessian stream network is a collection of 100 m long line segments, each containing a stream ID and a segment ID. The stream ID is designed in such a way, that it consecutively builds upon the ID of a previous stream segment, adding numbers to it, whenever the stream splits in upstream direction. The segment number increases towards the source of a stream (Fig. 2). Each 100 m segment already contains information on the in-stream and surrounding habitats structural quality (1 - pristine, 7 - completly altered).

```{r}
#| label: ID_example_figure
#| fig-cap: "Figure 2: Example of stream- and segment ID system. The segment number increases in the direction of the source. The stream ID of smaller confluence resembels always that of the bigger stream, in which it flows into, plus additional numbers."

str_net_plot <- stream_net %>% filter(str_starts(GEWKZ, "24848292")) %>% 
  ggplot()+
  geom_sf(
    aes(
      col = GEWKZ
    ),
    linewidth = 4
  )+
  geom_sf_text(
    aes(label = ABS)
    )+
  labs(col = "stream ID")+
  theme_bw()+
  theme(
    axis.title = element_blank(),
    legend.position = "bottom"
  )
str_net_plot
```

The `sfnetwork` package provides many useful functions for network analysis and routing operations, where the main function `as_sfnetwork()` builds a network by connecting aligning nodes. An initial build reveals that many stream segments visually seem connected, but are separated from their nearest node or edge by a few meters. Thus, to build a useful network for routing the precision of the stream network `sf` object needs to be lowered (for more information see Details on `st_set_precision()` and `st_as_binary()` help page). To what precision to round depends, in general, on the maximum distance of dangling nodes and if the network can be build by connecting nodes with edges, or just by connecting nodes. The precision will be set to 70 meters (be aware of the object CRS units!).

In the resulting network, the right streams are connected (Fig. 3) and it can be used for routing operations. Performance of routing can be greatly increased by simplifying the network and removing pseudo-nodes (more info [here](https://luukvdmeer.github.io/sfnetworks/articles/sfn02_preprocess_clean.html)).

``` r
# Lower precision to 50 meters
str_net_50m <- st_set_precision(stream_net, set_units(70, "m"))

# Create sf_network object, simplify and smooth pseudo-nodes
network <- as_sfnetwork(str_net_50m) %>% convert(to_spatial_simple) %>% 
  convert(to_spatial_smooth)
```

```{r}
#| label: Precision_example_figure
#| fig-cap: "Figure 3: Example for how precision changes the connectedness of the network. The numbers indicate from which node a stream originates."

net_1 <- stream_net %>% filter(str_starts(GEWKZ, "24848292")) %>% as_sfnetwork() %>% convert(to_spatial_simple) %>% convert(to_spatial_smooth)
net_2 <- stream_net %>% filter(str_starts(GEWKZ, "24848292")) %>% st_set_precision(0.1) %>% as_sfnetwork() %>% convert(to_spatial_simple) %>%
  convert(to_spatial_smooth)

edge_1 <- net_1 %>% st_as_sf("edges") %>% mutate(precision = "original precision")
edge_2 <- net_2 %>% st_as_sf("edges") %>% mutate(precision = "lowered precision")
nodes_1 <- net_1 %>% st_as_sf("nodes") %>% mutate(precision = "original precision")
nodes_2 <- net_2 %>% st_as_sf("nodes") %>% mutate(precision = "lowered precision")

edges <- rbind(edge_1, edge_2) %>% 
  mutate(precision = factor(precision, levels = c("original precision","lowered precision")))
nodes <- rbind(nodes_1, nodes_2) %>% 
  mutate(precision = factor(precision, levels = c("original precision","lowered precision")))

prec_plot <- edges %>% ggplot()+
  geom_sf(aes(color = as.factor(from)),
          linewidth = 2)+
  geom_sf(data = nodes, 
          color = "#000000", 
          size = 2) + 
  geom_sf_text(aes(label = from, 
                   color = as.factor(from)), 
               nudge_y = -100,
               nudge_x = -40,
               size = 7) + 
  facet_wrap(~precision, ncol = 2)+
  theme_bw()+ 
  theme(legend.position = "none", 
        axis.title = element_blank()
        )

prec_plot
```

Sampling sites and point stressors can be blend in as nodes into the created network.

``` r
# Blend in sampling sites, wastewater treatment plants and dams
network_blend <- st_network_blend(network, mzb) %>% st_network_blend(.,wwtp) %>% 
  st_network_blend(.,dams)
```

## Spatial Data quantification

For the quantification of point- and polygon features and their attributes I wrote two custom functions, `st_shift()` and `upstream_summarize()`. The first function is required within the latter and shifts a set of point geometries a proportional distance towards another specified point. The `upstream_summarize()` function builds a sub-network from a given point to all its vertices (stream sources) and extracts the node data, from which it counts the number of specified nodes and sums up specified attribute values. Further, it can calculate the minimal, maximal, or average network distance between the start point and a set of nodes specified by their IDs with the `IDs` argument. If no such nodes are present in the sub-network, `inf` or `NaN` values are returned. Additionally the Shreve stream magnitude, defined as the sum of sources upstream from a specifiedy node, can be calculated. The function can sum up attribute values of provided polygons the following way: First, the set of nodes present in the sub-network are shifted towards their centroid by 0.1% of their length, to avoid including adjacent polygons. Then, it creates a filtering mask by selecting all polygons which are touched by the shifted nodes and fills in 'holes' in the set of polygons. By setting a threshold, the mask can be shrunken to avoid selecting adjacent polygons. This mask is finally used to select all polygons present in the sub-network, from which their specified attributes are summarized. The function contains the following set of arguments:

-   `net` : The complete network with blended in points of interest
-   `start` : Row name of node from which to rout upstream
-   `node_cols` : Names of attribute columns to summarize, present in nodes
-   `IDs` : ID columns for sets of nodes, which should be unique in each set
-   `area` : `sf` object with only polygon geometries
-   `area_cols` : names of attribute columns to summarize, present in `area`
-   `dist` : Either `'min'`, `'max'`, `'mean'`, or `'all'`. Calculates distances between start and nodes, specified by `IDs`.
-   `threshold` : Value (in polygon CRS units) by which the polygon mask should be shrunken.
-   `Shreve` : logical, should Shreve stream magnitud be calculate?

Before this function can be used, the points from which the function should rout must be extracted as nodes from the network and their row name must be saved as a new column.

``` r
# Extract nodes of invertebrate sampling sites and add row names
mzb_nodes <- st_as_sf(network_blend, "nodes") %>% filter(!is.na(ID_SITE)) %>% 
  mutate(ID_NODE = row.names(nodes)[with(nodes, !is.na(ID_SITE))])
```

`upstream_summarize()` can be used in combination with rowwise() and mutate() to perform the action over multiple points. Depending on the number of nodes within the entire network, this task can take quite a lot of time. Luckily this task can be parallelized to save a lot of time.

``` r
# set up cluster
num_cores <- (detectCores()-1)
num_cores %>% makeCluster() %>% registerDoParallel()

# split data into chunks
data_chunks <- mzb_nodes %>% 
  split(., ceiling(seq_along(row_number(.)) / (length(row_number(.)) %/% num_cores)))

# Apply upstream_summarize parallel and row-wise over all sampling site nodes 
mzb_data_complete <- foreach(chunk = data_chunks, .combine = rbind, .packages = c("dplyr", "dtplyr","sf","sfnetworks","tidygraph","nngeo")) %dopar% {
  chunk %>% rowwise() %>% 
  mutate(upstream_summarize(
    net = network_blend,
    start = ID_NODE,
    node_cols = c("population_equivalents", "dam_discharge"),
    IDs = c("ID_WWTP", "ID_DAMS"),
    area = ws_clc,
    area_cols = c("Agriculture", "Urban", "semi-Natural"),
    dist = T,
    threshold = 30)
    )
}
```

## Data analysis and handling

### EDA

Before modeling it is good practice to get a sens of the target and explanatory variables. Here, multiple models should perform two different tasks, regression and multi-class classification. The respective target variables are the multi metric index (MMI) and the resulting ecological quality class (EQC). The MMI is a summarization of multiple core metrics, compared to reference conditions (for more details see [@hering2006]). It ranges from zero to one, representing bad to good condition respectively.

```{r}
#| label: MMI_histogram
#| fig-cap: "Figure 4: Histogram of the site averaged multi metric index."

data4model %>% ggplot()+
  geom_histogram(
    aes(x = MMI_mean),
    fill = "#99CCCC",
    colour = "#000000",
    bins = 10
    )+
    labs(
    title = "Multi metric index"
  )+
  theme_bw()+
  theme(
    axis.title.x = element_blank(),
    axis.title.y = element_text(size = 15, face = "bold"),
    axis.text = element_text(size = 12)
  )
```

```{r}
#| label: EQC_histogram
#| fig-cap: "Figure 5: Histogram of the site averaged ecological quality class."

data_all %>% ggplot()+
  geom_histogram(
    aes(x = EQC_mean),
    stat = "count",
    fill = "#FF99FF",
    colour = "#000000"
    )+
  labs(
    title = "Ecological quality class"
  )+
  theme_bw()+
  theme(
    axis.title.x = element_blank(),
    axis.title.y = element_text(size = 15, face = "bold"),
    axis.text = element_text(size = 12)
  )
```

From the EQC histogram (Fig. 5) a strong class imbalance is observable. To counter this and increase model performance, the data for the minority class could be oversampled using \[...\]. Since the classes are of ordered categorical, the minority could be merged with an adjacent class, although that would mean a loss of information.

Some covariates follow the same method of quantification and are therefore auto correlated by default. Sites at bigger reaches will have more point stresors in the upstream watershed compared to sites further upstream, therefore the different stressor covariates are automatically correlated. The correlation can be displayed in a correlation plot (Fig. 6).

```{r}
#| label: autocorrelation_plot
#| fig-cap: "Figure 6: Correlation plot of covariates."

data4model %>% select(TOT_WASTE, dam, num_ID_WWTP, num_SWOF_ID, cros_hw, cros_rail, dist_ID_WWTP_min, dist_SWOF_ID_min, Urban_prop:semi_Natural_prop) %>% 
  chart.Correlation(histogram = T, pch = 19)

```

Although tree ensemble learning algorithms are robust to auto-correlated covariates, it is best practice to either drop some related covariates or reduce the dimensionallity and use PCA scores as new variables.

### Spatial autocorrelation

Toblers first law states that *"everything is related to everything else, but near things are more related than distant things"*. This degree can be quantified by the Moranes Index, which can be calculated with distance-based weight matrix for points.

``` r
# k-nearest neighbours 
suppressPackageStartupMessages(require(deldir)) data_nb_knn <- knn2nb(knearneigh(data_all,k = 1))  
# distance neighbours 
dsts <- unlist(nbdists(data_nb_knn, data_all)) summary(dsts) max_1nn <- max(dsts)  data_nb_dst <- dnearneigh(data_all, d1 = 0, d2 = 0.75*max_1nn)  data_lw <- nb2listw(data_nb_dst, style = "W", zero.policy = T)  
# Moran statistics 
moran.test(data_all$MMI_mean, data_lw, alternative = "greater") moran.plot(data_all$MMI_mean, data_lw) 
```

```{r}
#| label: Moranes_index

suppressPackageStartupMessages(require(deldir))
data_nb_knn <- knn2nb(knearneigh(data_all,k = 1))

# distance neighbours
dsts <- unlist(nbdists(data_nb_knn, data_all))
summary(dsts)
max_1nn <- max(dsts)

data_nb_dst <- dnearneigh(data_all, d1 = 0, d2 = 0.75*max_1nn)

data_lw <- nb2listw(data_nb_dst, style = "W", zero.policy = T)

# Moran statistics
moran.test(data_all$MMI_mean, data_lw, alternative = "greater")
```

The index shows a significant deviation from the expectation, signifying strong spatial autocorrelation. This can also be visualized with a Moranes plot.

```{r}
#| label: moranes_plot
#| fig-cap: "Figure 6: Moranes plot for the site averaged multi metric index."

moran.plot(data_all$MMI_mean, data_lw, xlab = "mean MMI", ylab = "spatially lagged mean MMI")
```

Consequently, a geographically weighted XGBoost model should be calculated following [@li2019][@schimohr2022], by using the geo-referenced predictions from the XGBoost model as inputs for the geographically wheighted regression (GWR) model. Further, spatial heterogenity should be kept in mind for model tuning and cross validation [@schratz2019].

### Data cleaning & feature engeneering

The following variables were used to build the model:

Tree ensemble learning algorithms can handle covariates of different types and are robust to auto-correlated covariates. Nevertheless, it is good practice to clean and modify desired features before modeling. For each sampling site land use types in the upstream watershed were calculated as proportions to the total watershed area. Distance variables where transformed into relative inverse distances:

$$
dist_{inverse} = 1 - x / (x_{max} + 5 m)
$$

Where $x$ is the distance to the closest point stressor in meters and $x_{max}$ is the maximal distance to the closest point stressor in meters.

The Wastewater impact was calculated as the cumulative number of population equivalents from all WWTPs upstream of a sampling site, divided by the Shreve stream magnitude of the sampling site.

## Modeling

### Ensemble machine learning with XGBoost

We first separate the available data into a training and test set by a ratio of 70% to 30% respectively.

``` r
set.seed(1234)

# create split
inTrain <- createDataPartition(
  y = data4model$EQC_mean,
  p = 0.8,
  list = F
)

# split data
```

### GW-XGBoost

### Model performance

The performance of regression models can be evaluated by comparison of the mean squared errors. The classification models can be compared using the missclassification rate. The predicted values from the regression models can be translated to the EQC, which can also be evaluated via the misclassification rate.

``` r
```

The results in Table XX reveal that model XX is the best performing regression model while model XX is the best classifiaction model. Further, the geographically wheighted

## Data availability

Data on WFD invertebrate sampling and the Hessian stream network were kindly provided by the Hessian state office for nature, environment and geology (HLNUG). Data on wastewater treatment plants, stormwater overflows are available for download from the [WRRL-viewer](https://wrrl.hessen.de/mapapps/resources/apps/wrrl/index.html?lang=de). Further, data on hydrological barriers are available from the [Amber Barrier Atlas](https://amber.international/european-barrier-atlas/). OpenStreetMap data can be downloaded as pbf files at [geofabrik](https://download.geofabrik.de/). CORINE land cover shapefiles and rasters are available at [Copernicus Land Monitoring Service](https://land.copernicus.eu/en/products/corine-land-cover). The SRTM GL1 30m digital elevation model can be downloaded from [OpenTopography](https://portal.opentopography.org/raster?opentopoID=OTSRTM.082015.4326.1).

# References
